|

### 2020-08-11 pre-analysis-plan/freire-skarbek-lynching.Rmd
small edits in pre-analysis plan

[^thanks]: We thank Rosario Aguilar, Miriam Golden, Umberto Mignozzetti, Irfan Nooruddin, Catarina Roman, and Alex Sacco for their valuable comments.

@@

Why do citizens endorse extralegal punishment?[^definitions] Although mob
violence has decreased in advanced democracies, vigilantism remains widespread
in the developing world. In particular, Latin America has been severely affected
by a recent increase in lynchings. Motivated by growing numbers of drug-related
crime and by police ineffectiveness, citizen violence has surged in the region
[@mallen2014vigilantes]. Brazil provides a telling example. From 2011 to 2015,
the country registered about 2,500 lynching episodes, and in 2015 alone 173
people were killed by angry mobs, an average of one execution every two days
[@zechmeister2014political]. According to José Martins, who has studied
lynchings in Brazil for more than 30 years, those figures are not only the
highest in the country's history, but also the highest in the world
[@pearson2018latam]. Victims are generally accused of petty theft, and many are
beaten to death for stealing mobile phones, cooking utensils, or pairs of
sandals [@barbara2015vigilantes]. Lynching perpetrators are often young men,
but they also include female teenagers, elderly women, and even members of
local police forces [@moura2017linchamentos; @oliveira2016mob].

@@

Recent research suggests that support for vigilantism in Latin America is
associated with exposure to violence [@garcia2019anger], untrustworthy law
enforcement [@zizumbo2017community], perceived insecurity [@ceobanu2011crime;
@godoy2004justice], and social inequality [@arias2010violent]. However, approval
for extrajudicial violence raises important questions about the relationship
between states and societies in the region [@nivette2016institutional]. First,
while vigilantism provides social order to vulnerable populations, it also
undermines the legitimacy of state institutions [@schuberth2013challenging].
Lynching episodes reinforce the idea that civilians should not wait for legal
proceedings and that social control is better exerted by the people themselves
[@black1983crime]. Second, victim characteristics greatly impact the likelihood
and severity of lynchings. In Brazil, a person is rarely lynched when they
commit a crime against males, and females that commit the same crimes as males
also receive less punishment.[^fapesp] Moreover, there is no significant social
difference between victims and perpetrators, nor black Brazilians are
particularly targeted by lynch mobs [@oliveira2016mob]. These dynamics pose a
puzzle to existing explanations for mob violence support and stand in sharp
contrast with the American experience [@smaangs2016doing; @wood2011lynching].

@@

In this study, we endeavour to answer the following questions: 1) which victim
characteristics increase individual support for extralegal punishment?; 2) how
do individuals justify their approval for vigilantism?; and 3) can information
provision reduce support for mob violence? We design a set of three experiments
to tackle these issues. Our first experiment consists of a conjoint analysis
in which we show respondents different profiles of lynching victims and ask them
who they believe deserve the punishment. In our second experiment, we test how
ineffective policing, slow criminal justice systems, and demand for hasher legal
punishment affect individual propensity to support mob violence. Finally, we
assess whether citizens become less supportive of vigilantism when they are
informed about the legal consequences of lynchings, constitutional human rights,
or the risk of _vendettas_ associated with mob violence. While in this
pre-analysis plan we present the three experiments as described above,
their order will be randomised in our survey to mitigate potential order bias.

@@

\begin{table}[htpb]
\small
\begin{center}
\caption{\textbf{Attributes and Levels}}
\label{tab:categories}
\begin{tabular}{l !{\vrule width 1pt}p{9cm}}
\Xhline{2\arrayrulewidth}
\textbf{Attribute} & \multicolumn{1}{c}{\textbf{Levels}} \\
\Xhline{2\arrayrulewidth}
\small
Gender of crime perpetrator & Male; female \\ [4pt]
Age of crime perpetrator & Teenager; adult; elderly \\ [4pt]
Race of crime perpetrator & Black; white; native Brazilian; Asian \\ [4pt]
Offense & Picks the pocket; steals the car; molests; rapes; murders \\ [4pt]
Gender of crime victim & Male; female\\ [4pt]
Age of crime victim & Child; teenager; adult; elderly\\ [4pt]
Lynching perpetrators & Bystanders; neighbours; family of the victim; gangs; police \\
\Xhline{2\arrayrulewidth}
\end{tabular}
\end{center}
\end{table}

@@

Research shows that police ineffectiveness frequently appears as a strong
predictor for vigilantism [@cruz2019determinants; @garcia2019anger]. The direct
result of the weakness of police institutions is that citizens decide to take
criminal matters "in their own hands", thus persecuting and punishing the
criminals by themselves. A recent statistic indicates that the police
solves only 10% of the homicides in Brazil, what lends evidence to the link
between weak law enforcement and lynchings [@pearson2018latam].

Another determinant of support for vigilante justice that is suggested in the
literature is the lack of trust in the justice system [@godoy2004justice]. This
is often due to long criminal proceedings, which cause significant anxiety for
the victims. In Brazil, the penal code allows the accused to appeal each
decision several times, so it can take decades before a criminal case is closed
[@sousa2005utilizaccao]. In this respect, citizens do not believe that, even if
the criminal is put to trial, he/she will be punished in a timely matter. Note
that although the police is part of the criminal justice system, we separate the
two institutions in our experiment.

>>>>

[^thanks]: We thank Louise Araújo, Rosario Aguilar, Miriam Golden, Umberto Mignozzetti, Irfan Nooruddin, Catarina Roman, and Alex Sacco for their valuable comments.

@@

Why do citizens endorse extralegal punishment?[^definitions] Although mob
violence has decreased in advanced democracies, vigilantism remains widespread
in the developing world. In particular, Latin America has been severely affected
by a recent increase in lynchings. Motivated by growing numbers of drug-related
crime and by police ineffectiveness, citizen violence has surged in the region
[@mallen2014vigilantes]. Brazil provides a telling example. From 2011 to 2015,
the country registered about 2,500 lynching episodes, and in 2015 alone 173
people were killed by angry mobs, an average of one execution every two days
[@barbara2015vigilantes; @oliveira2016mob]. According to José Martins, who has
studied lynchings in Brazil for more than 30 years, those figures are not only
the highest in the country's history, but also the highest in the world
[@pearson2018latam]. Victims are generally accused of petty theft, and many are
beaten to death for stealing mobile phones, cooking utensils, or pairs of
sandals [@barbara2015vigilantes]. Lynching perpetrators are often young men,
but they also include female teenagers, elderly women, and even members of
local police forces [@moura2017linchamentos; @oliveira2016mob].

@@

Recent research suggests that support for vigilantism in Latin America is
associated with exposure to violence [@garcia2019anger], untrustworthy law
enforcement [@zizumbo2017community], perceived insecurity [@ceobanu2011crime;
@godoy2004justice], and social inequality [@arias2010violent]. However, approval
for extrajudicial violence raises important questions about the relationship
between states and societies in the region [@nivette2016institutional]. First,
while vigilantism provides social order to vulnerable populations, it also
undermines the legitimacy of state institutions [@schuberth2013challenging].
Lynching episodes reinforce the idea that civilians should not wait for legal
proceedings and that social control is better exerted by the people themselves
[@black1983crime]. Second, victim characteristics greatly impact the likelihood
and severity of lynchings. In Brazil, a person is rarely lynched when they
commit a crime against males, and females that commit the same crimes as males
also receive less punishment.[^fapesp] Moreover, there is no significant social
difference between victims and perpetrators, nor are black Brazilians 
particularly targeted by lynch mobs [@oliveira2016mob]. These dynamics pose a
puzzle to existing explanations for mob violence support and stand in sharp
contrast with the American experience [@smaangs2016doing; @wood2011lynching].

@@

In this study, we endeavour to answer the following questions: 1) which victim
characteristics increase individual support for extralegal punishment?; 2) how
do individuals justify their approval for vigilantism?; and 3) can information
provision reduce support for mob violence? We design a set of three experiments
to tackle these issues. Our first experiment consists of a conjoint analysis
in which we show respondents different profiles of lynching victims and ask them
who they believe deserve punishment. In our second experiment, we test how
ineffective policing, slow criminal justice systems, and demand for hasher legal
punishment affect individual propensity to support mob violence. Finally, we
assess whether citizens become less supportive of vigilantism when they are
informed about the legal consequences of lynchings, constitutional human rights,
or the risk of _vendettas_ associated with mob violence. While in this
pre-analysis plan we present the three experiments as described above,
their order will be randomised in our survey to mitigate potential order bias.

@@

\begin{table}[htpb]
\small
\begin{center}
\caption{\textbf{Attributes and Levels}}
\label{tab:categories}
\begin{tabular}{l !{\vrule width 1pt}p{9cm}}
\Xhline{2\arrayrulewidth}
\textbf{Attribute} & \multicolumn{1}{c}{\textbf{Levels}} \\
\Xhline{2\arrayrulewidth}
\small
Gender of crime perpetrator & Male; female \\ [4pt]
Age of crime perpetrator & Teenager; adult; elderly \\ [4pt]
Race of crime perpetrator & Black; White; Native Brazilian; Asian \\ [4pt]
Offense & Picks the pocket; steals the car; molests; rapes; murders \\ [4pt]
Gender of crime victim & Male; female\\ [4pt]
Age of crime victim & Child; teenager; adult; elderly\\ [4pt]
Lynching perpetrators & Bystanders; neighbours; family of the victim; gangs; police \\
\Xhline{2\arrayrulewidth}
\end{tabular}
\end{center}
\end{table}

@@

Research shows that police ineffectiveness frequently appears as a strong
predictor for vigilantism [@cruz2019determinants; @garcia2019anger]. The direct
result of the weakness of police institutions is that citizens decide to take
criminal matters "into their own hands", thus persecuting and punishing the
criminals by themselves. A recent statistic indicates that the police
solve only 10% of the homicides in Brazil, what lends support to the link
between weak law enforcement and lynchings [@pearson2018latam].

Another determinant of support for vigilante justice that is suggested in the
literature is the lack of trust in the justice system [@godoy2004justice;
@smith2019contradictions]. This is often due to long criminal proceedings,
which cause significant anxiety for the victims. In Brazil, the penal code
allows the accused to appeal each decision several times, so it can take
decades before a criminal case is closed [@sousa2005utilizaccao]. In this
respect, citizens do not believe that, even if the criminal is put to trial,
he/she will be punished in a timely matter. Note that although the police is
part of the criminal justice system, we separate the two institutions in our
experiment.
